---
title: "GrandRiver22:microbiome_alphadiversity"
author: "Milena Esser"
date: "2025-05-20"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

# Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      results = "hide")
rm(list=ls()) #Clears environment
#Change margin size to be smaller so graphs fit in the plot panel
par(mar = c(2, 2, 2, 2)) # Set the margin on all sides to 2
#setwd("C:\\Users\\milen\\OneDrive - McMaster University\\Bow river microbiome\\R_analysis")

###Loading packages
library(phyloseq) #phyloseq will be the main package used for structuring microbiome data and diversity comparisons
library(ggplot2) #For creating graphs
library(ggtext)
library(plyr) #if you load this after dpylr it messes up the dplyr functions
library(dplyr) #Helps with data wrangling 
library(vegan)
library(tidyverse) #data wrangling
library(knitr) #For R Markdown knitting
library(RColorBrewer)
library(ggpubr)
library(MicEco) #to prune based on prevalence
library(picante)
library(ggh4x) # For facet_nested_wrap()
library(multcompView)
library(microbiome) #abundances function Retrieves the taxon abundance table from phyloseq-class object and ensures it is systematically returned as taxa x samples matrix.
library(rdacca.hp) #package for hierarchical and variation partitioning in multiple regression and canonical ananlysis
library(data.table)
```

## Loading Data

Loading the phyloseq objects

```{r Loading data}
#not rarefied dataset "Bow_ps"
load("GrandRiverMB_phyloseq_filt_unnorm.RData")

#rarefied dataset "ps_rare"
load("GrandRiverMB_phyloseq_filt_rar.RData")

#css-normalized dataset "ps_css"
load("GrandRiverMB_phyloseq_filt_css.RData")

#TSS-normalized dataset "ps_prop"
load("GrandRiverMB_phyloseq_filt_propnorm.RData")
```

#### Select data set for analysis

```{r}
dataset <- "css"
ps <- get(paste0("ps_", dataset))

# Create output directory
output_dir <- paste0("R_output//Beta_diversity//", dataset) 
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## Beta diversity

```{r Beta diversity}
#### Redundancy analysis (RDA) ALL SAMPLES ####

ps_bray<-phyloseq::distance(ps, method="bray") #convert to distance matrix

t_otu_all <- t(abundances(ps_bray)) #transpose distance matrix
metadata_all <- meta(ps) #get sample data

metadata_all

#Only looking at site and Family
predictors <- metadata_all[, c("Site_ID", "Sample_Type")]


rdacca.hp(t_otu_all, predictors, method="RDA", type = "adjR2") #Run the RDA to see the percent explained by each factor
#quantify how much each predictor (or group of predictors) uniquely and jointly explains the variation in your response data.


```

### PCoA Bray Curtis - ALL SAMPLES

```{r}
#ordination
bray <- ordinate(
  physeq = ps, 
  method = "PCoA", 
  distance = "bray" 
)


#Scree plot of PCoA
scree <- plot_scree(bray)
scree + theme(axis.text.x=element_text(size=8, color="black"), axis.text.y=element_text(size=18, color="black"), 
        axis.text=element_text(size=18, color="black"), axis.title=element_text(size=18)) 
#axis 1, 2, and 3 account for most variation explained (mostly axis 1)

metadata<- data.frame(sample_data(ps))

metadata$Site_ID <- factor(metadata$Site_ID)
metadata$Sample_Type <- factor(metadata$Sample_Type)

#All together
plot_ordination(ps, bray, shape="Site_ID", color="Sample_Type") + #Plots the ordination
  theme_classic() +
  theme(strip.background = element_blank())

#save plot
filename <- file.path(output_dir, paste0("PCoA_Bray_ALLsamples_", dataset, ".png"))
ggsave(filename = filename, height = 5, width = 7)

#Faceted by stage (larvae, adults, spiders)
plot_ordination(ps, bray, shape="Site_ID", color="Sample_Type") + 
  theme_classic() +
  theme(strip.background = element_blank())+
  facet_wrap(~Sample_Type)



```

### NMDS Bray Curtis -ALL SAMPLES

```{r}
# NMDS 
set.seed(501)
ord_NMDS = ordinate(ps, "NMDS", "bray", trymax=100, k=3) 
ord_NMDS #< 0.05 = excellent, 0.05–0.1 = very good, 0.1–0.2 = usable but some distortion, > 0.2 = not reliable
ord_NMDS = ordinate(ps, "NMDS", "bray", trymax=100, k=2) 
ord_NMDS #< 0.05 = excellent, 0.05–0.1 = very good, 0.1–0.2 = usable but some distortion, > 0.2 = not reliable

#Weak ties mean that few identical distances were tied during ranking, which is good — it indicates the data have good variability.

#Test wether 3D may be better k=3 than k=2


#Plotting all sites and taxa in one facet
plotNMDS1<-plot_ordination(ps, ord_NMDS, shape="Site_ID", color = "Sample_Type") + 
  theme_classic() +
  geom_point(size=3)+
  theme(strip.background = element_blank())+
  theme(legend.key.size = unit(0.3, 'cm'), 
        legend.title = element_text(size=20), 
        legend.text = element_text(size=18), 
        axis.text = element_text(size=16, color="black"),
        axis.title = element_text(size=18, color="black"),
        legend.position = "right")+
  guides(color=guide_legend(title="Invertebrate Taxa")) +
  scale_color_discrete(palette(c("orange","blue","red","black","magenta","peachpuff")))


newtab1 = data.table(plotNMDS1$data)
newtab1$Family <- ordered(newtab1$Family,
 levels=c("aeshnidae", "heptageniidae", "chironomidae", "zygoptera", "araneidae", "tetragnathidae"), labels=c("Aeshnidae", "Heptageniidae", "Chironomidae", "Zygoptera", "Araneidae", "Tetragnathidae"))
plotNMDS1$data <- newtab1
print(plotNMDS1)

#save plot
filename <- file.path(output_dir, paste0("NMDS_Bray_ALLsamples_", dataset, ".png"))
ggsave(filename = filename, height = 10, width = 14)



#Ellipses by taxa
plotNMDStaxa<-plot_ordination(ps, ord_NMDS, color = "Sample_Type", shape="Site_ID") + 
  theme_classic() +
  geom_point(size=3)+
  theme(strip.background = element_blank())+
  theme(legend.key.size = unit(0.3, 'cm'), 
        legend.title = element_text(size=20), 
        legend.text = element_text(size=18), 
        axis.text = element_text(size=16, color="black"),
        axis.title = element_text(size=18, color="black"),
        legend.position = "right")+
  guides(color=guide_legend(title="Invertebrate Taxa")) +
  scale_color_discrete(palette(c("orange","blue","red","black","magenta","peachpuff")))+
  stat_ellipse(type="norm", alpha=1, aes(group=Sample_Type, color=Sample_Type))

#newtabtaxa = data.table(plotNMDStaxa$data)
#newtabtaxa$taxgroup <- ordered(newtabtaxa$taxgroup,
#levels=c("aeshnidae", "heptageniidae", "chironomidae", "zygoptera", "araneidae", "tetragnathidae"), labels=c("Aeshnidae", "Heptageniidae", "Chironomidae", "Zygoptera", "Araneidae", "Tetragnathidae"))
#plotNMDStaxa$data <- newtabtaxa
print(plotNMDStaxa)

#save plot
filename <- file.path(output_dir, paste0("NMDS_Bray_ALLsamples_elipses_", dataset, ".png"))
ggsave(filename = filename, height = 10, width = 14)


#Ellipses by Site
plotNMDSsite<-plot_ordination(ps, ord_NMDS, color = "Sample_Type", shape="Site_ID") + 
  theme_classic() +
  geom_point(size=3)+
  theme(strip.background = element_blank())+
  theme(legend.key.size = unit(0.3, 'cm'), 
        legend.title = element_text(size=20), 
        legend.text = element_text(size=18), 
        axis.text = element_text(size=16, color="black"),
        axis.title = element_text(size=18, color="black"),
        legend.position = "right")+
  guides(color=guide_legend(title="Invertebrate Taxa")) +
  scale_color_discrete(palette(c("orange","blue","red","black","magenta","peachpuff")))+
  stat_ellipse(type="norm", alpha=1, aes(group=Site_ID))

#newtabtaxa = data.table(plotNMDSsite$data)
#newtabtaxa$taxgroup <- ordered(newtabtaxa$taxgroup,
#levels=c("aeshnidae", "heptageniidae", "chironomidae", "zygoptera", "araneidae", "tetragnathidae"), labels=c("Aeshnidae", "Heptageniidae", "Chironomidae", "Zygoptera", "Araneidae", "Tetragnathidae"))
#plotNMDSsite$data <- newtabtaxa
print(plotNMDSsite)

#save plot
filename <- file.path(output_dir, paste0("NMDS_Bray_ALLsamples_EllipsesSITE_", dataset, ".png"))
ggsave(filename = filename, height = 10, width = 14)



```

### Statistics

#### PERMANOVA

```{r}
##Create Bray curtis distance matrix
df_bray <- phyloseq::distance(ps, method = "bray")

## PERMANOVA by site and invertebrate taxa ##

test.adonis.all <- adonis2(df_bray ~ Site_ID*Sample_Type, data = metadata)
test.adonis.all




```

#### Beta dispersion

betadisper measures the spread (i.e., variability) of samples around the group centroid in NMDS (or distance) space. It’s analogous to a Levene’s test for homogeneity of variance, but for multivariate data. It helps to understand if your PERMANOVA results are driven more by location (centroid differences) or spread (dispersion differences).

If dispersions differ too much, you have to be cautious in interpreting PERMANOVA results. As this is highly significant: Differences in microbial composition among Families are not only due to changes in the mean community but also because some Families are more variable than others.

-   **If p \> 0.05**: No evidence of unequal dispersion → You can trust group separation in PERMANOVA.

<!-- -->

-   **If p \< 0.05**: Significant dispersion → Group differences in PERMANOVA **may be confounded** by group variance, **not just location**.

```{r}

#Are microbial (or community) compositions more variable in some Families than in others?
bd_all_fam<-betadisper(df_bray, metadata$Sample_Type) 
anova(bd_all_fam)


#### Boxplot to show dispersion #####
# Extract distances to centroids
beta_df <- data.frame(
  Family = bd_all_fam$group,
  DistanceToCentroid = bd_all_fam$distances
)

# Quick look at the structure
head(beta_df)

```

In case of p\< 0.05: pairwise comparison to identify group with significantly different dispersion

```{r}
tukey_res <- TukeyHSD(bd_all_fam)
sig_disp <- as.data.frame(tukey_res$group)
sig_disp$Comparison <- rownames(sig_disp)
sig_disp <- sig_disp[sig_disp$`p adj` < 0.05, ]

# View the significant ones
sig_disp[order(sig_disp$`p adj`), ]
```

Create boxplot

```{r}
# Create a ggplot boxplot
ggplot(beta_df, aes(x = Family, y = DistanceToCentroid, fill = Family)) +
  geom_boxplot(alpha = 0.7, outlier.color = "black", outlier.size = 1.5) +
  geom_jitter(width = 0.2, alpha=0.5, size=1.5) + # add jittered points
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size=12),
    axis.text.y = element_text(size=12),
    axis.title = element_text(size=14, face="bold"),
    legend.position = "none"
  ) +
  labs(
    x = "Family",
    y = "Distance to Group Centroid",
 #   title = "Beta Dispersion across Invertebrate Families"
  ) +
  scale_fill_brewer(palette = "Set2")  # optional color palette

#Save plots

#Save plot as png
filename <- file.path(output_dir, paste0("Beta_Dispersion_across_Invertebrate_Families_", dataset, ".png"))
ggsave(filename = filename, height = 5, width = 8)
#save plot as pdf
filename <- file.path(output_dir, paste0("Beta_Dispersion_across_Invertebrate_Families_", dataset, ".pdf"))
ggsave(filename = filename, height = 5, width = 8)


#Beta dispersion differed significantly among Families, suggesting that observed differences in community composition are associated with both shifts in mean composition and differences in within-group variability.


bd_all_site<-betadisper(df_bray, metadata$Site_ID) 
anova(bd_all_site) 
```

### NMDS Bray Curtis - IND. TAXA (Adult vs Larvae)

```{r}
# Define your list of Taxa
unique_taxa <- unique(sample_data(ps)$Taxa)

# Initialize result containers
nmds_stress_values <- numeric()
nmds_results <- list()
permanova_results <- list()
betadispersion_results <- list()

# Define color palette
lifestage_colors <- c("Adult" = "#1f78b4", "Larvae" = "#33a02c")
location_colors <- c("Upstream" = "#e31a1c", "Downstream" = "#ff7f00")

# Loop over each Taxa
for (taxon in unique_taxa) {
  cat("\nProcessing Taxa:", taxon, "\n")

  # Subset data by Taxa
  ps_sub <- subset_samples(ps, Taxa == taxon)
  ps_sub <- prune_samples(sample_sums(ps_sub) > 0, ps_sub)

  if (nsamples(ps_sub) < 3) {
    cat("Skipping", taxon, "- not enough samples.\n")
    next
  }

  # Check if Lifestage exists
  metadata <- as(sample_data(ps_sub), "data.frame")
  has_lifestage <- "Lifestage" %in% colnames(metadata) && length(unique(metadata$Lifestage)) > 1

  # Distance matrix and ordination
  dist <- phyloseq::distance(ps_sub, method = "bray")
  set.seed(501)
  ord <- tryCatch(ordinate(ps_sub, method = "NMDS", distance = dist, trymax = 100), error = function(e) NULL)

  if (is.null(ord) || is.null(ord$points)) {
    cat("NMDS failed for", taxon, "- skipping plot.\n")
    next
  }

  # Save stress value
  stress_val <- round(ord$stress, 3)
  cat("NMDS Stress for", taxon, ":", stress_val, "\n")
  nmds_stress_values[taxon] <- stress_val
  nmds_results[[taxon]] <- ord

  # PERMANOVA
  permanova_site <- adonis2(dist ~ Site_ID, data = metadata)
  permanova_location <- adonis2(dist ~ Location, data = metadata)
  permanova_results[[taxon]] <- list(Site = permanova_site, Location = permanova_location)
  cat("PERMANOVA Site_ID R2:", round(permanova_site$R2[1], 3), "p =", permanova_site$`Pr(>F)`[1], "\n")
  cat("PERMANOVA Location R2:", round(permanova_location$R2[1], 3), "p =", permanova_location$`Pr(>F)`[1], "\n")

  # Beta Dispersion
  beta_site <- betadisper(dist, metadata$Site_ID)
  beta_location <- betadisper(dist, metadata$Location)
  beta_anova_site <- anova(beta_site)
  beta_anova_location <- anova(beta_location)
  betadispersion_results[[taxon]] <- list(Site = beta_anova_site, Location = beta_anova_location)
  cat("Betadispersion Site_ID p-value:", beta_anova_site$`Pr(>F)`[1], "\n")
  cat("Betadispersion Location p-value:", beta_anova_location$`Pr(>F)`[1], "\n")

  # PERMANOVA and Beta Dispersion for Lifestage (if available)
  if (has_lifestage) {
    permanova_lifestage <- adonis2(dist ~ Lifestage, data = metadata)
    beta_lifestage <- betadisper(dist, metadata$Lifestage)
    beta_anova_lifestage <- anova(beta_lifestage)

    permanova_results[[taxon]]$Lifestage <- permanova_lifestage
    betadispersion_results[[taxon]]$Lifestage <- beta_anova_lifestage

    cat("PERMANOVA Lifestage R2:", round(permanova_lifestage$R2[1], 3), "p =", permanova_lifestage$`Pr(>F)`[1], "\n")
    cat("Betadispersion Lifestage p-value:", beta_anova_lifestage$`Pr(>F)`[1], "\n")
  }

  # Plot NMDS
  color_var <- if (has_lifestage) "Lifestage" else "Location"
  color_scale <- if (has_lifestage) scale_color_manual(values = lifestage_colors) else scale_color_manual(values = location_colors)

  plot <- plot_ordination(ps_sub, ord, type = "samples") +
    geom_point(aes_string(color = color_var, shape = "Location"), size = 3) +
    color_scale +
    theme_classic() +
    labs(title = paste("NMDS -", taxon), color = color_var, shape = "Location")

  # Save plot
  ggsave(file.path(output_dir, paste0("NMDS_", taxon, "_", dataset, "_Bray.png")), plot = plot, width = 8, height = 6, dpi = 300)
  ggsave(file.path(output_dir, paste0("NMDS_", taxon, "_", dataset, "_Bray.pdf")), plot = plot, width = 8, height = 6, dpi = 300)
  cat("Saved NMDS plot for", taxon, "\n")
}


```

Looking up individual results and running Stats

```{r}
summary_table <- data.frame(
  Taxa = character(),
  Permanova_Site_R2 = numeric(),
  Permanova_Site_p = numeric(),
  Permanova_Location_R2 = numeric(),
  Permanova_Location_p = numeric(),
  Permanova_Lifestage_R2 = numeric(),
  Permanova_Lifestage_p = numeric(),
  Betadispersion_Site_p = numeric(),
  Betadispersion_Location_p = numeric(),
  Betadispersion_Lifestage_p = numeric(),
  NMDS_Stress = numeric(),
  stringsAsFactors = FALSE
)

for (taxon in unique_taxa) {
  if (!is.null(permanova_results[[taxon]]) && !is.null(betadispersion_results[[taxon]])) {
    permanova <- permanova_results[[taxon]]
    beta_disp <- betadispersion_results[[taxon]]

    # Defaults if Lifestage not tested
    perm_lifestage_r2 <- NA
    perm_lifestage_p <- NA
    beta_lifestage_p <- NA

    if (!is.null(permanova$Lifestage)) {
      perm_lifestage_r2 <- round(permanova$Lifestage$R2[1], 3)
      perm_lifestage_p <- round(permanova$Lifestage$`Pr(>F)`[1], 4)
    }

    if (!is.null(beta_disp$Lifestage)) {
      beta_lifestage_p <- round(beta_disp$Lifestage$`Pr(>F)`[1], 4)
    }

    summary_table <- rbind(summary_table, data.frame(
      Taxa = taxon,
      Permanova_Site_R2 = round(permanova$Site$R2[1], 3),
      Permanova_Site_p = round(permanova$Site$`Pr(>F)`[1], 4),
      Permanova_Location_R2 = round(permanova$Location$R2[1], 3),
      Permanova_Location_p = round(permanova$Location$`Pr(>F)`[1], 4),
      Permanova_Lifestage_R2 = perm_lifestage_r2,
      Permanova_Lifestage_p = perm_lifestage_p,
      Betadispersion_Site_p = round(beta_disp$Site$`Pr(>F)`[1], 4),
      Betadispersion_Location_p = round(beta_disp$Location$`Pr(>F)`[1], 4),
      Betadispersion_Lifestage_p = beta_lifestage_p,
      NMDS_Stress = ifelse(!is.na(nmds_stress_values[taxon]), nmds_stress_values[taxon], NA)
    ))
  }
}

# Save summary table to CSV
print(summary_table)
write.csv(summary_table, file = file.path(output_dir, paste0("NMDS_Summary_Stats_taxa_", dataset, ".csv")), row.names = FALSE)
cat("Summary table saved as 'NMDS_Summary_Stats_taxa_", dataset, ".csv'\n")

# Save stress values as a separate CSV
write.csv(as.data.frame(nmds_stress_values), file = file.path(output_dir, paste0("NMDS_Stress_Values_", dataset, ".csv")))
cat("NMDS stress values saved as 'NMDS_Stress_Values_", dataset, ".csv'\n")
```

### NMDS Bray Curtis - IND. TAXA and LIFESTAGES

```{r}
# Define your list of Sample_Types
unique_sample_type <- unique(sample_data(ps)$Sample_Type)

# Initialize result containers
nmds_stress_values <- numeric()
nmds_results <- list()
permanova_results <- list()
betadispersion_results <- list()

# Define color palette
location_colors <- c("UP" = "#009900", "DWN" = "#e31a1c")

# Loop over each Sample_Type
for (sample_type in unique_sample_type) {
  cat("\nProcessing Sample_Type:", sample_type, "\n")

  # Subset data by Sample_Type
  ps_sub <- subset_samples(ps, Sample_Type == sample_type)
  ps_sub <- prune_samples(sample_sums(ps_sub) > 0, ps_sub)

  if (nsamples(ps_sub) < 3) {
    cat("Skipping", sample_type, "- not enough samples.\n")
    next
  }

  # Get corresponding metadata
  metadata_sub <- data.frame(sample_data(ps_sub))

  # Distance matrix and ordination
  dist <- phyloseq::distance(ps_sub, method = "bray")
  set.seed(501)
  ord <- tryCatch(ordinate(ps_sub, method = "NMDS", distance = dist, trymax = 100), error = function(e) NULL)

  if (is.null(ord) || is.null(ord$points)) {
    cat("NMDS failed for", sample_type, "- skipping plot.\n")
    next
  }

  # Save stress value
  stress_val <- round(ord$stress, 3)
  cat("NMDS Stress for", sample_type, ":", stress_val, "\n")
  nmds_stress_values[sample_type] <- stress_val
  nmds_results[[sample_type]] <- ord

  # PERMANOVA
  permanova_site <- adonis2(dist ~ Site_ID, data = metadata_sub)
  permanova_location <- adonis2(dist ~ Location, data = metadata_sub)
  permanova_results[[sample_type]] <- list(Site = permanova_site, Location = permanova_location)

  cat("PERMANOVA Site_ID R2:", round(permanova_site$R2[1], 3), "p =", permanova_site$`Pr(>F)`[1], "\n")
  cat("PERMANOVA Location R2:", round(permanova_location$R2[1], 3), "p =", permanova_location$`Pr(>F)`[1], "\n")

  # Beta Dispersion
  beta_site <- betadisper(dist, metadata_sub$Site_ID)
  beta_location <- betadisper(dist, metadata_sub$Location)
  beta_anova_site <- anova(beta_site)
  beta_anova_location <- anova(beta_location)

  betadispersion_results[[sample_type]] <- list(Site = beta_anova_site, Location = beta_anova_location)

  cat("Betadispersion Site_ID p-value:", beta_anova_site$`Pr(>F)`[1], "\n")
  cat("Betadispersion Location p-value:", beta_anova_location$`Pr(>F)`[1], "\n")

  # Plot NMDS
  plot <- plot_ordination(ps_sub, ord, type = "samples") +
    geom_point(aes(color = Location, shape = Site), size = 3) +
    scale_color_manual(values = location_colors) +
    theme_classic() +
    labs(title = paste("NMDS -", sample_type), color = "Location", shape = "Location")

  # Save plots
  ggsave(file.path(output_dir, paste0("NMDS_", sample_type, "_", dataset, "_Bray.png")), plot = plot, width = 8, height = 6, dpi = 300)
  ggsave(file.path(output_dir, paste0("NMDS_", sample_type, "_", dataset, "_Bray.pdf")), plot = plot, width = 8, height = 6, dpi = 300)
  cat("Saved NMDS plot for", sample_type, "\n")
}

```

Summary stats

```{r}
summary_table <- data.frame(
  Sample_Type = character(),
  Permanova_Site_R2 = numeric(),
  Permanova_Site_p = numeric(),
  Permanova_Location_R2 = numeric(),
  Permanova_Location_p = numeric(),
  Betadispersion_Site_p = numeric(),
  Betadispersion_Location_p = numeric(),
  NMDS_Stress = numeric(),
  stringsAsFactors = FALSE
)

for (sample_type in unique_sample_type) {
  if (!is.null(permanova_results[[sample_type]]) && !is.null(betadispersion_results[[sample_type]])) {
    permanova <- permanova_results[[sample_type]]
    beta_disp <- betadispersion_results[[sample_type]]

    summary_table <- rbind(summary_table, data.frame(
      Sample_Type = sample_type,
      Permanova_Site_R2 = round(permanova$Site$R2[1], 3),
      Permanova_Site_p = round(permanova$Site$`Pr(>F)`[1], 4),
      Permanova_Location_R2 = round(permanova$Location$R2[1], 3),
      Permanova_Location_p = round(permanova$Location$`Pr(>F)`[1], 4),
      Betadispersion_Site_p = round(beta_disp$Site$`Pr(>F)`[1], 4),
      Betadispersion_Location_p = round(beta_disp$Location$`Pr(>F)`[1], 4),
      NMDS_Stress = ifelse(!is.na(nmds_stress_values[sample_type]), nmds_stress_values[sample_type], NA)
    ))
  }
}

# Save summary table to CSV
print(summary_table)
write.csv(summary_table, file = file.path(output_dir, paste0("NMDS_Summary_Stats_sampletype_", dataset, ".csv")), row.names = FALSE)
cat("Summary table saved as 'NMDS_Summary_Stats_sampletype_", dataset, ".csv'\n")

# Save stress values as a separate CSV
write.csv(as.data.frame(nmds_stress_values), file = file.path(output_dir, paste0("NMDS_Stress_Values_", dataset, ".csv")))
cat("NMDS stress values saved as 'NMDS_Stress_Values_", dataset, ".csv'\n")

```

### Identifying outlier

There seems to be an outlier in Heptageniidae Adults with MDS1 \> 4000

```{r, eval = FALSE}
# Define the sample_type you're interested in
target_sample_type <- "Adult_Heptageniidae"  # Replace with your actual sample_type name

# Extract NMDS results from the list
ord <- nmds_results[[target_sample_type]]

  # Extract NMDS coordinates
  nmds_points <- as.data.frame(ord$points)
  nmds_points$SampleID <- rownames(nmds_points)

  # Identify samples with extreme NMDS1 values
  outliers <- nmds_points[nmds_points$MDS1 > 4000, ]

#identified KK2580 as outlier
```

identified KK2580 (WAT_DWN3_Ad_Mayfly_MB_3) as outlier (will sort out in first script (creating phyloseq object) and rerun this analysis
